Generalized spaces for constructive algebra

### 2020-04-08 paper-herrsching2019/main.tex
Proper formatting of the case studies

\paragraph{Applications in constructive algebra.}
Commutative algebra abounds with techniques to reduce given situations to
simpler ones, for instance passing to a quotient or passing to a
localization. These techniques facilitate short and elegant proofs, such as the
following. (Section~\ref{sect:algebraic-preliminaries} contains some algebraic
preliminaries.)

@@

\begin{example}For future reference, and also for giving a different
perspective on statement~(\ref{item:negneg-stable}), we unroll here the
interpretation of~$f \models \neg\neg\varphi$. It is:
\begin{quote}
For any element~$g \in A$, \\
${\qquad}$ if for any element~$h \in A$, \\
${\qquad\qquad}$ if~$fgh \models \varphi$, \\
${\qquad\qquad}$ then~$fgh$ is nilpotent, \\
${\qquad}$ then~$fg$ is nilpotent.
\end{quote}
In the case that~$A$ is reduced and~$f$ is the unit element of~$A$, this
condition implies (by setting~$g \defeq 1$): ``If~$h = 0$ is the only element of~$A$
such that~$h \models \varphi$, then~$1 = 0$ in~$A$.''
\end{example}

@@

\begin{quote}
\textbf{Theorem.} Let~$M$ be an injective matrix with more columns than rows
over a reduced ring~$A$. Then~$1 = 0$ in~$A$.
\end{quote}

@@

\begin{quote}
Over the localized
ring~$A[M_{ij}^{-1}]$, the matrix element~$M_{ij}$ is invertible and hence~$M$ can be
put into the form~$\left(\begin{smallmatrix}1&0\\0&M'\end{smallmatrix}\right)$
by elementary row and column transformations. The matrix~$M'$ is injective and
has more columns than rows, hence by induction, applied to this submatrix
over~$A[M_{ij}^{-1}]$, we have~$1 = 0$ in~$A[M_{ij}^{-1}]$. Since~$A$ is
reduced, this amounts to~$M_{ij} = 0$ in~$A$.

Hence~$M$ is the zero matrix. Hence~$M(1,0,\ldots,0)^t = 0$ and by
injectivity~$1 = 0$ in~$A$.
\end{quote}

@@

\paragraph{Grothendieck's generic freeness theorem.}

\begin{quote}
\textbf{Theorem.}
Let~$A$ be a reduced ring. Let~$M$ be a finitely generated~$A$-module.
If~$f = 0$ is the only element of~$A$ such that~$M[f^{-1}]$ is a
free~$A[f^{-1}]$-module, then~$1 = 0$ in~$A$.
\end{quote}

\begin{quote}
\textbf{Proof.} The claim amounts to the statement that~$M^\sim$ is \notnot
free. Since~$A^\sim$ is a field, this statement follows from
basic (intuitionistic) linear algebra.\footnote{A standard result of linear
algebra is: ``Any finitely generated vector space over a field is finite free
(that is, has a finite basis).'' The proof proceeds by considering a given
generating family and using the law of excluded middle to determine whether one
of the generators can be expressed using the others or not. In the latter case,
the family is linearly independent and hence a basis; in the former case, the
proof removes the redundant generator and continues by induction.

Intuitionistically, we only have the double negation of the law of excluded
middle available; this is why intuitionistically we only have that finitely
generated vector spaces are \notnot finite free. The field condition required
by the proof is exactly the one satisfied by~$A^\sim$.}\qed\end{quote}

Again, the proof can be unwound if desired:

@@

We finish by using the assumption for~$f = 1$. \qed
\end{quote}

\begin{quote}
\textbf{Theorem.}
Let~$A$ be a reduced ring. Let~$B$ be an~$A$-algebra of finite type. Let~$M$ be
a finitely generated~$B$-module.
If~$f = 0$ is the only element of~$A$ such that
\begin{enumerate}
\item $B[f^{-1}]$ and $M[f^{-1}]$ are free modules over $A[f^{-1}]$,
\item $A[f^{-1}] \to B[f^{-1}]$ is of finite presentation and
\item $M[f^{-1}]$ is finitely presented as a module over~$B[f^{-1}]$,
\end{enumerate}
then~$1 = 0$ in~$A$.
\end{quote}

\begin{quote}\textbf{Proof.} One can check that the claim is the interpretation by the sheaf
semantics of the following statement about the sheaves: It is \notnot the case that
\begin{enumerate}
  \item $B^\sim$ and $M^\sim$ are free modules over $A^\sim$ (by basic linear algebra),
  \item $A^\sim \to B^\sim$ is of finite presentation, and
  \item $M^\sim$ is finitely presented as a module over $B^\sim$,
\end{enumerate}

Item~(1) is by basic linear algebra as in the proof of
simple version of Grothendieck's generic freeness lemma, but a bit more involved as we now need to
deal with countable generating
families~\cite[Theorem~11.16]{blechschmidt:phd}. Item~(2) is because finitely
generated algebras over anonymously Noetherian rings are \notnot finitely
presented (in a description of the form~$R[X_1,\ldots,X_n]/\aaa$, the
ideal~$\aaa$ is \notnot finitely generated).
Item~(3) is because finitely generated modules over anonymously Noetherian are
\notnot finitely presented.\qed
\end{quote}

A variant of this proof where the reference to the Noetherian condition has
been replaced by an explicit computation has been unwound in
Ref.~\refcite{blechschmidt:generic-freeness}.

@@

\begin{quote}
\textbf{Proof.} Let~$x \in A$ be such that~$x (\Lambda^m M) = (0)$. We are to
show~$x = 0$. By passing to the factor ring~$A/\ann(x)$, we may
assume~$(\Lambda^m M) = (0)$ and~$x = 1$; the new goal is to verify~$1 = 0$.

@@

Since~$M$ is injective, the ideal generated by the first column is regular,
hence under~$\nabla$ the unit ideal. Since our goal is~$\nabla$-stable, we may assume
that it actually is the unit ideal. Hence one of the entries in the first
column is invertible. Applying elementary row and column transformations, we
can put~$M$ into the
form~$\left(\begin{smallmatrix}1&0\\0&M'\end{smallmatrix}\right)$. Continuing
in this fashion, we obtain the
form~$\left(\begin{smallmatrix}I\\0\end{smallmatrix}\right)$. This matrix
obviously has an invertible~$m$-minor, and hence the original one has one as
well.
\end{quote}
\end{alphlist}
\end{exercise}

\begin{exercise}{Bridging modules and their induced sheaves}%
\label{ex:bridging-modules}%
Let~$M$ be a module over a ring~$A$.
\begin{alphlist}[(c)]
\item Show that~$M$ is finitely generated iff~$M^\sim$ is.
\item Show that~$M$ is finite locally free (that is, that there is a
partition~$1 = f_1 + \cdots + f_n$ such that each module~$M[f_i^{-1}]$ is
finite free over~$A[f_i^{-1}]$) iff~$M^\sim$ is finite free.
\item Show that~$M$ is finitely presented iff~$M^\sim$ is.
\end{alphlist}
\end{exercise}

>>>>

\paragraph{Applications in constructive algebra.}
\label{par:appl-constr-alg}
Commutative algebra abounds with techniques to reduce given situations to
simpler ones, for instance passing to a quotient or passing to a
localization. These techniques facilitate short and elegant proofs, such as the
following. (Section~\ref{sect:algebraic-preliminaries} contains some algebraic
preliminaries.)

@@

\begin{example}\label{ex:negneg-interpretation}
For future reference, and also for giving a different
perspective on statement~(\ref{item:negneg-stable}), we unroll here the
interpretation of~$f \models \neg\neg\varphi$. It is:
\begin{quote}
For any element~$g \in A$, \\
${\qquad}$ if for any element~$h \in A$, \\
${\qquad\qquad}$ if~$fgh \models \varphi$, \\
${\qquad\qquad}$ then~$fgh$ is nilpotent, \\
${\qquad}$ then~$fg$ is nilpotent.
\end{quote}
In the case that~$A$ is reduced and~$f$ is the unit element of~$A$, this
condition implies (by setting~$g \defeq 1$): ``If~$h = 0$ is the only element of~$A$
such that~$h \models \varphi$, then~$1 = 0$ in~$A$.''
\end{example}

@@

On page~\ref{par:appl-constr-alg}, we presented a classical proof of the
following theorem.
\begin{quote}
\textbf{Theorem.} Let~$M$ be an injective matrix with more columns than rows
over a reduced ring~$A$. Then~$1 = 0$ in~$A$.
\end{quote}
We are now in a position to show how the reduction technique developed in
Section~\ref{sect:appl} can be used to recast the classical proof so that it
becomes constructive:

@@

\begin{quote}
\textbf{Proof.} Over the localized
ring~$A[M_{ij}^{-1}]$, the matrix element~$M_{ij}$ is invertible and hence~$M$ can be
put into the form~$\left(\begin{smallmatrix}1&0\\0&M'\end{smallmatrix}\right)$
by elementary row and column transformations. The matrix~$M'$ is injective and
has more columns than rows, hence by induction, applied to this submatrix
over~$A[M_{ij}^{-1}]$, we have~$1 = 0$ in~$A[M_{ij}^{-1}]$. Since~$A$ is
reduced, this amounts to~$M_{ij} = 0$ in~$A$.

Hence~$M$ is the zero matrix. Thus~$M(1,0,\ldots,0)^t = 0$ and by
injectivity~$1 = 0$ in~$A$. \qed
\end{quote}

@@

\paragraph{Grothendieck's generic freeness lemma.} This theorem is a basic
theorem in algebraic geometry, used for instance for developing the theory of
moduli spaces. When phrased for the affine situation in a constructively
sensible way, a simple version of this theorem reads as follows.
\begin{quote}
\textbf{Theorem.}
Let~$A$ be a reduced ring. Let~$M$ be a finitely generated~$A$-module.
If~$f = 0$ is the only element of~$A$ such that~$M[f^{-1}]$ is a
free~$A[f^{-1}]$-module, then~$1 = 0$ in~$A$.
\end{quote}
This theorem is particularly interesting for the program of constructivizing
classical commutative algebra, as it is an example where the constructive proof
is much shorter and arguably more perspiciuous than classical proofs (for which
see for instance Refs.~\cite[Lemme~6.9.2]{ega-4-2},
\cite[Thm.~24.1]{matsumura:commutative-ring-theory} or
\cite[Thm.~14.4]{eisenbud:commutative-algebra}, which only cover the case of
Noetherian integral domains, or Refs.~[\refcite{staats:generic-freeness}],
\cite[{\href{https://stacks.math.columbia.edu/tag/051Q}{Tag~051Q}}]{stacks-project}
which proceed in a series of intermediate steps, reducing to that case).
\begin{quote}
\textbf{Proof.} By Example~\ref{ex:negneg-interpretation} and
Exercise~\ref{ex:bridging-modules}(\ref{item:free}), the claim amounts to the statement that~$M^\sim$ is \notnot
free. Since~$A^\sim$ is a field, this statement follows from
basic (intuitionistic) linear algebra.\footnote{A standard result of linear
algebra is: ``Any finitely generated vector space over a field is finite free
(that is, has a finite basis).'' The proof proceeds by considering a given
generating family and using the law of excluded middle to determine whether one
of the generators can be expressed using the others or not. In the latter case,
the family is linearly independent and hence a basis; in the former case, the
proof removes the redundant generator and continues by induction.

Intuitionistically, we only have the double negation of the law of excluded
middle available; this is why intuitionistically we only have that finitely
generated vector spaces are \notnot finite free. The field condition required
by the proof is exactly the one satisfied by~$A^\sim$.}\qed\end{quote}
Again, the proof can be unwound if
desired~\cite[Proposition~3]{blechschmidt:generic-freeness}:

@@

We finish by using the assumption for~$f = 1$. \qed
\end{quote}
The full version of Grothendieck's generic freeness lemma, and a constructive
proof of it, read as follows.
\begin{quote}
\textbf{Theorem.}
Let~$A$ be a reduced ring. Let~$B$ be an~$A$-algebra of finite type. Let~$M$ be
a finitely generated~$B$-module.
If~$f = 0$ is the only element of~$A$ such that
\begin{enumerate}
\item $B[f^{-1}]$ and $M[f^{-1}]$ are free modules over $A[f^{-1}]$,
\item $A[f^{-1}] \to B[f^{-1}]$ is of finite presentation and
\item $M[f^{-1}]$ is finitely presented as a module over~$B[f^{-1}]$,
\end{enumerate}
then~$1 = 0$ in~$A$.

\textbf{Proof.} One can check that the claim is the interpretation by the sheaf
semantics of the following statement about the sheaves: It is \notnot the case that
\begin{enumerate}
  \item $B^\sim$ and $M^\sim$ are free modules over $A^\sim$ (by basic linear algebra),
  \item $A^\sim \to B^\sim$ is of finite presentation, and
  \item $M^\sim$ is finitely presented as a module over $B^\sim$,
\end{enumerate}

Item~(1) is by basic linear algebra as in the proof of
simple version of Grothendieck's generic freeness lemma, but a bit more involved as we now need to
deal with countable generating
families~\cite[Theorem~11.16]{blechschmidt:phd}. Item~(2) is because finitely
generated algebras over anonymously Noetherian rings are \notnot finitely
presented (in a description of the form~$R[X_1,\ldots,X_n]/\aaa$, the
ideal~$\aaa$ is \notnot finitely generated).
Item~(3) is because finitely generated modules over anonymously Noetherian are
\notnot finitely presented.\qed
\end{quote}
A variant of this proof where the reference to the Noetherian condition has
been replaced by an explicit computation has been unwound in
Ref.~\refcite{blechschmidt:generic-freeness}.

@@

\begin{quote}
\textbf{Proof of the theorem.} Let~$x \in A$ be such that~$x (\Lambda^m M) = (0)$. We are to
show~$x = 0$. By passing to the factor ring~$A/\ann(x)$, we may
assume~$(\Lambda^m M) = (0)$ and~$x = 1$; the new goal is to verify~$1 = 0$.

@@

Since~$M$ is injective, the ideal generated by the first column is regular,
hence under~$\nabla$ the unit ideal. Since our goal is~$\nabla$-stable, we may assume
that it actually is the unit ideal. Hence one of the entries in the first
column is invertible. Applying elementary row and column transformations, we
can put~$M$ into the
form~$\left(\begin{smallmatrix}1&0\\0&M'\end{smallmatrix}\right)$. Continuing
in this fashion, we obtain the
form~$\left(\begin{smallmatrix}I\\0\end{smallmatrix}\right)$. This matrix
obviously has an invertible~$m$-minor, and hence the original one has one as
well.\qed
\end{quote}
\end{alphlist}
\end{exercise}

\begin{exercise}{Bridging modules and their induced sheaves}%
\label{ex:bridging-modules}%
Let~$M$ be a module over a ring~$A$.
\begin{alphlist}[(c)]
\item Show that~$M$ is finitely generated iff~$M^\sim$ is.
\item\label{item:free} Show that~$M$ is finite locally free (that is, that there is a
partition~$1 = f_1 + \cdots + f_n$ such that each module~$M[f_i^{-1}]$ is
finite free over~$A[f_i^{-1}]$) iff~$M^\sim$ is finite free.
\item Show that~$M$ is finitely presented iff~$M^\sim$ is.
\end{alphlist}
\end{exercise}
